The imprint of global magma oceans on exoplanet demographics

### 2023-03-30 src/tex/ms.tex
outline story for introduction

\begin{note}
For observations of rocky exoplanets, the currently best-probed regime is that of warm, close-in planets.
These bodies experience climatic conditions that are similar to the environment of the inner Solar System bodies at early stages of their formation.
Studies of the geophysical state and evolution of hot exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.

...
    Venus and Earth, while having accreted from the same mass reservoir and despite their similar bulk properties, evolved into planets with very different environmental conditions on their surfaces.
    At the formation of the Solar System, both planets underwent a giant impact phase [CITE!] that melted their mantles, leading to a magma ocean stage.
    Due to these similar early formation phases, it was commonly assumed that the divergence of Venus and Earth -- in particular Venus's water loss -- occurred late in their evolution~\citep{Elkins-Tanton2013}.
    \citet{Hamano2013} showed that such a late desiccation process is not needed to explain the stark differences: for a planet rich in volatiles and receiving high enough radiation levels from its host star, a steam atmosphere limits the outgoing radiation from the (molten) planetary surface.
    This runaway greenhouse state prevents the planet's rapid cooling and can extend the magma ocean stage to hundreds of \SI{}{\mega\year}, enough to remove the entire water reservoir from a rocky planet by hydrodynamic escape.
    \todo[inline]{important for motivating the existence of the pattern: ``the solidification timescale can become comparable to the main-sequence lifetime of the star (Hamano et al. 2013, 2015)'' \citep{Lichtenberg2022} }
    Consequently, it is aptly considered to be the inner boundary of the habitable zone~\citep{Kasting1993,Kopparapu2013}.

    (Under the right conditions, different orbital distances alone can thus decide about a planet's water content and ultimately its habitability.)
    If received flux makes the difference between our own habitable planet and dry, dead Venus, there is good reason to believe that such patterns exist in exoplanet systems as well.

    ...long-term goal/overarching objective: derive the geophysical history of rocky extrasolar planets. More concrete: Constrain the limits of runaway greenhouse transitions, and thus the inner edges of the habitable zone. Also: How close is Earth from this runaway greenhouse irradiation limit?
   ...
\todo[inline]{introduce magma oceans and their influence on planetary radii~\citep{Dorn2021}.}
    ...current/future observations of planets that are currently in the runaway greenhouse phase may constrain properties of their planetary mantles... make connection between interior and atmospheres.
    ...

    The runaway greenhouse transition is a robust prediction from climate models, and its impact on planetary bulk properties has been shown to be in the detectable range of current astronomical instrumentation~\citep{Goldblatt2015}.
Measuring these impacts would indeed provide the first empirical evidence of the habitable zone concept.
On the planetary level, however, such undertaking is hindered by degeneracies: From internal composition to atmospheric loss to volatile content, many factors influence a planet's size.
It is therefore prudent to shift the focus to the planet population as a whole and search for a statistical effect in exoplanet demographics~\citep{Turbet2019}.
 Using \bioverse\ to combine the predictions from geophysical models of steam atmospheres with a synthesis of planets informed by observed demographics, we...

    Planetary radius changes due to the combined effect of steam atmospheres and water dissolved in the molten mantle of planets within the runaway greenhouse threshold is expected to cause a discontinuity in the radius distribution of small (DEFINITION) exoplanets.

\end{note}

@@

\section{Discussion}\label{sec:discussion}
\todo[inline]{\textbf{Potential additional discussion points: }\\
(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}

@@

\subsubsection{Detection in multi-planet systems}
As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five systems.
Future additions to the multi-planet sample through missions such as \plato\ are needed.

>>>>

\begin{itemize}
    \item hot/warm planets are best studied
    \item runaway greenhouse climate (Ingersoll1969,Kasting1988,Nakajima1992)
    \item HZ concept.''(Kasting et al. 1993; Kopparapu et al. 2013, 2014)'' Relevance in the search for habitable planets and life. (Science gap:) Does it exist? Can it be tested? => statistically detect inner edge
    \item Bioverse
    \item Transit survey => PLATO
    \item Using PLATO as a baseline survey, we do A, B, C.
\end{itemize}

%\begin{note}
%Recent years have brought tremendous advances in our understanding of terrestrial-sized planets, mainly due to space-based transit surveys such as \kepler\ \citep{Borucki2010} and \tess\ \citep{Ricker2014a} that pushed detection sensitivities into the terrestrial regime.
%Due to the inherent detection biases of the transit technique, most current well-probed small planets are much closer to their host star than Earth is to the Sun. CITE
%These bodies experience climatic conditions that are similar to the environment of the inner Solar System planets at early stages of their formation. CITE
%Studies of the geophysical state and evolution of warm exoplanets can thus contribute to our understanding of the early formation stages of Earth and other habitable worlds.
%
%A much regarded object of study are the divergent atmospheric evolutions of Earth and Venus.
%While both have accreted from the same mass reservoir and despite their similar bulk properties, they evolved into planets with very different environmental conditions on their surfaces. "(Donahue et al. 1982; Kasting 1988; Hamano et al. 2013; Way et al. 2016; Kane et al. 2019; Way \& Del Genio 2020; Turbet et al. 2021)."
%
%
%
%
%
%
%
%
%
%\end{note}

@@

\section{Discussion}\label{sec:discussion}
%\todo[inline]{\textbf{Potential additional discussion points: }\\
%(How close is Earth to runaway? (-> \citet{Goldblatt2012,Popp2016}; what is the max. expected change in Earth's albedo due to deglaciation?)}

@@

\subsubsection{Detection in multi-planet systems}
As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.
The current sample of suitable systems is sparse: The California-Kepler Survey catalog~\citep{Fulton2018} contains only six planets with instellations $< 2\, S_\oplus$ and smaller than \SI{2}{\rEarth} in five multi-planet systems.
Future additions to the multi-planet sample through missions such as \plato\ are needed.
